Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-18 prop_letter.tex
prop_letter.tex Reviewing...

A principal result is that recasting the Kochen-Specker theorem in
terms of ``$\delta$-determinism'', with~$\delta$ quantifying the
effect of finite-precision uncertainties on measurement outcomes,
reducing to the purely deterministic Kochen-Specker case for
$\delta=0$, can resolve the question of the effect of finite precision
and imperfections on contextuality.  A remarkable consequence of our
analysis is that not only does the quantum-mechanical requirement for
contextuality continue to hold, thus maintaining the Kochen-Specker
result for a \emph{non-vanishing} range of $\delta$, but that at a
\emph{certain fixed value} of $\delta$, there is a sharp transition,
parallel in its consequences to, say, a phase transition, to
\emph{non-contextuality}; that is, suddenly a quantum system can be
simulated by a non-contextual hidden variable theory, turning the
system effectively from a quantum system into a classical system. 
% that is, quantum mechanics suddenly fails to hold, and all variables
% can be specified simultaneously, turning the theory effectively into a
% classical theory that can be specified consistently by hidden
% variables.
% \yutsung{The part after ``that is''
%   looks strange.
% \begin{enumerate}
% \item The Kochen-Specker theorem is not about quantum mechanics, is about whether
% quantum mechanics can be simulated by a non-contextual deterministic hidden
% variable theory or not. Hence, ``quantum mechanics suddenly fails to hold''
% doesn't make a lot sense.
% \item The question is whether there exist a QIVPM means ``a classical theory
% consistently by hidden variables.'' At least, a QIVPM is not deterministic, and
% I need to check whether it qualities as a classical probabilistic hidden variable
% theory\ldots(?)
% \end{enumerate}
% }
These results provide a new insight into the Meyer-Mermin debate by
presenting a theory in which there is a \emph{parameter} interpolating
between what appeared previously to be irreconcilable interpretations
of contextuality.

@@

As an example, consider a three-dimensional Hilbert space with orthonormal
basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable $\mathbf{O}$
with spectral decomposition $\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$,
i.e., $\lambda_{i}=j$ and $P_{i}=\proj{j-1}$. Two fragments of valid
probability measures~$\mu_{1}$ and $\mu_{2}$ that can be associated
with this space are defined in Table~\ref{tab:quantum-probability-measure}.
\begin{table}
\noindent \centering{}\caption{\label{tab:quantum-probability-measure}Two fragments of valid probability
measures~$\mu_{1}$ and $\mu_{2}$.}
\begin{tabular}{ccc}
\toprule 
\addlinespace
$\ket{\psi}$ & $\mu_{1}(\proj{\psi})$ & $\mu_{2}(\proj{\psi})$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\ket{0}$ & $\frac{1}{2}$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{1}$ & $\frac{1}{2}$ & $0$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{2}$ & $0$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}$ & $1$ & $\frac{1}{4}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\rmi}=\frac{\ket{0}+\rmi\ket{1}}{\sqrt{2}}$ & $\frac{1}{2}$ & $\frac{1}{4}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\ps'}=\frac{\ket{0}+\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\rmi'}=\frac{\ket{0}+\rmi\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\ps''}=\frac{\ket{1}+\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{4}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\rmi''}=\frac{\ket{1}+\rmi\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{4}$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By the Born rule, the first probability measure corresponds to the
quantum system being in the pure state $\ket{\ps}=(\ket{0}+\ket{1})/\sqrt{2}$
and the second corresponds to the quantum system being in the state
$\rho=(\proj{0}+\proj{2})/2$. The expectation values of the observable
$\mathbf{O}$, $\expval{\mathbf{O}}_{\mu_{1,2}}$, are 1.5 in the
first case and 2 in the second.

@@

\noindent The definition puts no restrictions on the choice of intervals,
only on the assignment of intervals to events. When $\delta=0$, every
event must be assigned a probability either in the interval $\imposs$
or in $\necess$, i.e., every event is completely determined with
certainty. As $\delta$ gets larger, the QIVPM allows for more indeterminate
behavior. To illustrate the example easier, we restrict ourselves
to the classical subspace of commuting events~$\events_{0}=\left\{ \mathbb{0},\proj{0},\proj{1},\mathbb{1}\right\} $.
If we apply the same definition in the classical case, the classical
IVPM~$\bar{\mu}_{4}$ defined in Table~\ref{tab:classical-IVPMs}
is $0$-deterministic because $\imposs\subseteq\left[0,0\right]$
and $\necess\subseteq\left[1-0,1\right]$.
\begin{table}
\noindent \centering{}\caption{\label{tab:classical-IVPMs}Classical IVPMs~$\bar{\mu}_{4}$, $\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ on $\events_{0}$.}
\begin{tabular}{cccc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}_{4}(P)$ & $\bar{\mu}_{5}(P)$ & $\bar{\mu}_{6}(P)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\necess$ & $\left[\frac{1}{3},1\right]$ & $\left[\frac{2}{3},\frac{2}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\imposs$ & $\left[0,\frac{1}{3}\right]$ & $\left[\frac{1}{3},\frac{1}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $\necess$ & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By interpreting the states ~$\proj{0}$ and $\proj{1}$ with the
head and the tail of a coin again, $\bar{\mu}_{4}$ represents tossing
a double-headed coin which can be predicted with certain. When $\delta$
is increased to $\frac{1}{3}$, the classical IVPMs~$\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ defined in Table~\ref{tab:classical-IVPMs}
are clearly $\frac{1}{3}$-deterministic, and they have more indeterminate
behavior than $\bar{\mu}_{4}$. However, the most indeterminate classical
IVPM~$\bar{\mu}_{2}$ defined in Eq.~(\ref{eq:classical-no-information-IVPM})
is still not $\frac{1}{3}$-deterministic because $\bar{\mu}_{2}(\proj{0})=\bar{\mu}_{2}(\proj{1})=\unknown$,
but neither $\unknown\subseteq\left[0,\frac{1}{3}\right]$ nor $\unknown\subseteq\left[1-\frac{1}{3},1\right]$.

@@

While the finite-precision extension of Kochen-Specker theorem asserts
the non-existence of certain kind of QIVPMs, we also want to understand
what a QIVPM could be when it exists. In Sec.~\ref{sec:Interval-Uncertainty},
the only non-trivial QIVPM~ $\bar{\mu}_{3}$ is induced by the state~$\proj{\ps}$.
The way we extending a state to a QIVPM will first be generalized
in this section, and introduce a finite-precision extension of the
Born rule~\citep{peres1995quantum,544199,Jaeger2007}. The Born rule
is essential to compute the conventional quantum probability because
any quantum probability measure can be induced by the Born rule according
to Gleason's theorem~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
We then provide a finite precision variant of Gleason's theorem for
the Born QIVPMs, but there are also some general QIVPMs may not be
induced by any quantum probability measure, i.e., no state is consistent
with these kind of QIVPMs.

>>>>

A principal result is that recasting the Kochen-Specker theorem in
terms of ``$\delta$-determinism'', with~$\delta$ quantifying the
effect of finite-precision uncertainties on measurement outcomes,
reducing to the purely deterministic Kochen-Specker case for
$\delta=0$, can resolve the question of the effect of finite precision
and imperfections on contextuality.  A remarkable consequence of our
analysis is that not only does the quantum-mechanical requirement for
contextuality continue to hold, thus maintaining the Kochen-Specker
result for a \emph{non-vanishing} range of $\delta$, but that at a
\emph{certain fixed value} of $\delta$, there is a sharp transition,
parallel in its consequences to, say, a phase transition, to
\emph{non-contextuality}; that is, suddenly a quantum system can be
simulated by a non-contextual hidden variable theory, turning the
system effectively from a quantum system into a classical system. 
% that is, quantum mechanics suddenly fails to hold, and all variables
% can be specified simultaneously, turning the theory effectively into a
% classical theory that can be specified consistently by hidden
% variables.
% \yutsung{The part after ``that is''
%   looks strange.
% \begin{enumerate}
% \item The Kochen-Specker theorem is not about quantum mechanics, is about whether
% quantum mechanics can be simulated by a non-contextual deterministic hidden
% variable theory or not. Hence, ``quantum mechanics suddenly fails to hold''
% doesn't make a lot sense.
% \item The question is whether there exist a QIVPM means ``a classical theory
% consistently by hidden variables.'' At least, a QIVPM is not deterministic, and
% I need to check whether it qualities as a classical probabilistic hidden variable
% theory\ldots(?)
% \end{enumerate}
% }
\yutsung{Still not comfortable about the current statement...}
These results provide a new insight into the Meyer-Mermin debate by
presenting a theory in which there is a \emph{parameter} interpolating
between what appeared previously to be irreconcilable interpretations
of contextuality.

@@

As an example, consider a three-dimensional Hilbert space with orthonormal
basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable $\mathbf{O}$
with spectral decomposition $\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$,
i.e., $\lambda_{i}=i$ and $P_{i}=\proj{i-1}$. Two fragments of valid
probability measures~$\mu_{1}$ and $\mu_{2}$ that can be associated
with this space are defined in Table~\ref{tab:quantum-probability-measure}.
\begin{table}
\noindent \centering{}\caption{\label{tab:quantum-probability-measure}Two fragments of valid probability
measures~$\mu_{1}$ and $\mu_{2}$.}
\begin{tabular}{ccc}
\toprule 
\addlinespace
$\ket{\psi}$ & $\mu_{1}(\proj{\psi})$ & $\mu_{2}(\proj{\psi})$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\ket{0}$ & $\frac{1}{2}$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{1}$ & $\frac{1}{2}$ & $0$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{2}$ & $0$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}$ & $1$ & $\frac{1}{4}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\rmi}=\frac{\ket{0}+\rmi\ket{1}}{\sqrt{2}}$ & $\frac{1}{2}$ & $\frac{1}{4}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\ps'}=\frac{\ket{0}+\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\rmi'}=\frac{\ket{0}+\rmi\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{2}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\ps''}=\frac{\ket{1}+\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{4}$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\ket{\rmi''}=\frac{\ket{1}+\rmi\ket{2}}{\sqrt{2}}$ & $\frac{1}{4}$ & $\frac{1}{4}$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By the Born rule, the first probability measure corresponds to the
quantum system being in the pure state $\ket{\ps}=(\ket{0}+\ket{1})/\sqrt{2}$
and the second corresponds to the quantum system being in the state
$\rho=(\proj{0}+\proj{2})/2$. The expectation values of the observable
$\mathbf{O}$, $\expval{\mathbf{O}}_{\mu_{1,2}}$, are 1.5 in the
first case and 2 in the second.

@@

\noindent The definition puts no restrictions on the choice of intervals,
only on the assignment of intervals to events. When $\delta=0$, every
event must be assigned a probability either in the interval $\imposs$
or in $\necess$, i.e., every event is completely determined with
certainty. As $\delta$ gets larger, the QIVPM allows for more indeterminate
behavior. To illustrate the example easier, we restrict ourselves
to the classical subspace of commuting events~$\events_{0}=\left\{ \mathbb{0},\proj{0},\proj{1},\mathbb{1}\right\} $.
If we apply the same definition in the classical case, the classical
IVPM~$\bar{\mu}_{4}$ defined in Table~\ref{tab:classical-IVPMs}
is $0$-deterministic because $\imposs\subseteq\left[0,0\right]$
and $\necess\subseteq\left[1-0,1\right]$.
\begin{table}
\noindent \centering{}\caption{\label{tab:classical-IVPMs}Classical IVPMs~$\bar{\mu}_{4}$, $\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ on $\events_{0}$.}
\begin{tabular}{cccc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}_{4}(P)$ & $\bar{\mu}_{5}(P)$ & $\bar{\mu}_{6}(P)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\necess$ & $\left[\frac{1}{3},1\right]$ & $\left[\frac{2}{3},\frac{2}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\imposs$ & $\left[0,\frac{1}{3}\right]$ & $\left[\frac{1}{3},\frac{1}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $\necess$ & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By interpreting the states ~$\proj{0}$ and $\proj{1}$ with the
head and the tail of a coin again, $\bar{\mu}_{4}$ represents tossing
a double-headed coin which can be predicted with certain. When $\delta$
is increased to $\frac{1}{3}$, the classical IVPMs~$\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ defined in Table~\ref{tab:classical-IVPMs}
are clearly $\frac{1}{3}$-deterministic, and they have more indeterminate
behavior than $\bar{\mu}_{4}$. However, the most indeterminate classical
IVPM~$\bar{\mu}_{2}$ defined in Eq.~(\ref{eq:classical-no-information-IVPM})
is still not $\frac{1}{3}$-deterministic because $\bar{\mu}_{2}(\proj{0})=\bar{\mu}_{2}(\proj{1})=\unknown$,
but neither $\unknown\subseteq\left[0,\frac{1}{3}\right]$ nor $\unknown\subseteq\left[1-\frac{1}{3},1\right]$.
Finally, when $\delta=1$, because every interval is a subset of $\unknown$,
every QIVPM is $1$-deterministic. In particular, the informativeness
QIVPM~$\bar{\mu}_{1}$ and the Born QIVPM~$\bar{\mu}_{3}$ defined
in Sec.~\ref{sec:Interval-Uncertainty} are both $1$-deterministic.

@@

While the finite-precision extension of Kochen-Specker theorem asserts
the non-existence of certain kind of QIVPMs, we also want to understand
what a QIVPM could be when it exists. In Sec.~\ref{sec:Interval-Uncertainty},
the only non-trivial QIVPM~ $\bar{\mu}_{3}$ is induced by the state~$\proj{\ps}$.
The way we extending a state to a QIVPM will first be generalized
in this section, and introduce a finite-precision extension of the
Born rule~\citep{Born1983,peres1995quantum,544199,Jaeger2007}. The
Born rule is essential to compute the conventional quantum probability
because any quantum probability measure can be induced by the Born
rule according to Gleason's theorem~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
We then provide a finite precision variant of Gleason's theorem for
the Born QIVPMs, but there are also some general QIVPMs may not be
induced by any quantum probability measure, i.e., no state is consistent
with these kind of QIVPMs.
